---
title: "R Notebook - Figure 6"
output: html_document
---

Required packages.
```{r, echo=TRUE}
library(readxl)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(msigdbr)
library(ggtext)
library(cowplot)
library(ggpubr)
```

Setup working directory.
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/omics/odcf/analysis/OE0285_projects/chip-in-bm/CK-AML_manuscript/') # Change to your working directory
```

## Plot Figure 6e and Extended Data Figure 10c UMAP plots

Read in single cell metadata object. 
```{r}
df.AML <- readxl::read_excel('./data/D1922_R0836_metadata.xlsx')

## Column IDs
# wnnUMAP_1: WNN-based UMAP co-ordiante (Dimension 1)
# wnnUMAP_2: WNN-based UMAP co-ordiante (Dimension 2)
# Cell: cell ID
# orig.ident: sample
# nCount_RNA: per cell RNA counts
# nFeature_RNA: per cell RNA features
# nCount_ADT: per cell ADT counts
# nFeature_ADT: per cell ADT features
# sample: sample ID
# labels: singleR-based annotation of a cell using healthy BM cells as reference (see Methods)
# subclone: subclone of the cell
# Ng_LSC_Up: Mean expression of the normalized gene counts of the signature genes obtained by Ng and colleagues (see Methods)
# polyclonal.id: growth pattern of the sample
# sample_subclone: combined sample and subclone information
# subclone_label: short subclone ID

```

Define labels, colors and shapes for plots.
```{r}
limits_paired =  c('Subclone1', 'Subclone2', 'Subclone3', 'Subclone4', 'Subclone5')
labels_paired = c('SC1', 'SC2', 'SC3', 'SC4', 'SC5')
values_paired = c("#4393c3","#91bfdb","#d6604d","#f4a582","#fdb863")
shape_paired = c(16,15)
sample_paired = c('D1922','R0836')
col_pair <- c("#ffb8b1","#993441")
names(col_pair) <- c('Diagnosis','Relapse')

myeloid.cells <- c("HSC","LMPP","ProgRBC", "ProgMk", "GMP","CD14 Mono","CD16 Mono","ProgDC","pDC","cDC2")
```

Plot UMAPs.
```{r}
# UMAP of diagnosis vs relapse myeloid cells
p_umap1 <- df.AML %>%
  filter(labels %in% myeloid.cells) %>%
  sample_frac() %>%
  ggplot(., aes(wnnUMAP_1, wnnUMAP_2)) +
  ggrastr::geom_point_rast(aes(color = stage),
                           size = 0.5, stroke = 0, raster.dpi = 500) +
  scale_color_manual(values = col_pair) +
  theme_pubr(legend = 'right') + 
  theme(legend.text = element_text(size = 12),
        legend.title = element_blank(), 
        axis.text=element_text(size=12),
        axis.title=element_text(size=14),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.ticks=element_blank(),
        panel.grid = element_blank(),
        plot.title = element_markdown()
  ) +
  labs(title = 'P5 diagnosis and relapse') +
  guides(color = guide_legend(override.aes = list(size=3))) +
  coord_fixed() +
  labs(x = "UMAP 1", y = "UMAP 2")

p_umap1

cowplot::save_plot(file = './figures/Figure6e_P5_UMAP_longitudinal.pdf', p_umap1, base_height = 4.5)

# UMAP of subclones
df.AML$subclone_sample <- paste0(df.AML$sample, '_', df.AML$subclone)

p_umap2 <- df.AML %>%
  filter(labels %in% myeloid.cells) %>%
  sample_frac() %>%
  ggplot(., aes(wnnUMAP_1, wnnUMAP_2)) +
  ggrastr::geom_point_rast(data = . %>% filter(subclone == 'Other'), color = 'lightgrey',
                           size = 0.5, stroke = 0, raster.dpi = 500) +
  ggrastr::geom_point_rast(data = . %>% filter(subclone != 'Other'), aes(color = subclone, shape = sample),
                           size = 0.75, stroke = 0, raster.dpi = 500) +
  scale_color_manual(values = values_paired,
                     limits = limits_paired,
                     labels = labels_paired) +
  scale_shape_manual(values = shape_paired,
                     limits = sample_paired) +
  theme_pubr(legend = 'right') + 
  theme(legend.text = element_text(size = 12),
        legend.title = element_blank(), 
        axis.text=element_text(size=12),
        axis.title=element_text(size=14),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.ticks=element_blank(),
        panel.grid = element_blank(),
        plot.title = element_markdown()
  ) +
  labs(title = 'P5 subclones at diagnosis and relapse') +
  guides(color = guide_legend(override.aes = list(size=3)), shape = guide_legend(override.aes = list(size=3))) +
  coord_fixed() +
  labs(x = "UMAP 1", y = "UMAP 2")

p_umap2

cowplot::save_plot(file = './figures/Extended10c_P5_UMAP_subclones.pdf', plot = p_umap2, base_height = 4.5)

```

## Plot Extended Data Figure 10b violin plots

Plot stemness scores.
```{r}
comparisons <- list(c('Diagnosis','Relapse'))

p_violin <- df.AML %>%
  filter(labels %in% myeloid.cells) %>%
  mutate(sample = factor(stage, levels = c('Diagnosis','Relapse'))) %>%
  ggplot(., aes(x = stage, y = Ng_LSC_Up)) +
  ggrastr::geom_quasirandom_rast(aes(color = stage), size = 0.1, dodge.width = 0.8) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange", colour = "black", size = 0.5) + # 95% confidence interval for the mean
  scale_color_manual(values = col_pair) +
  geom_signif(comparisons = comparisons, step_increase = 0.1, tip_length = 0.01, textsize = 3, margin_top = 0.05, color = 'black') +
  theme_pubr(legend = 'none') +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 0, vjust = 0.5),
        strip.background = element_blank()) +
  guides(fill = guide_legend(override.aes = list(size=3), title = "CLR ADT")) +
  labs(x = NULL, y = "Stemness Score")
  
p_violin

cowplot::save_plot(file = "./figures/Extended10b_P5_Ng_LSC_signature.pdf", plot = p_violin, base_width = 3, ncol = 1)
```

## Plot Figure 6f UMAP

Plot UMAPs and violin plots of interesting genes.
```{r}
df.int <- readxl::read_excel('./data/D1922_R0836_RNA_intgene.xlsx')

int.genes <- c('SOX4','EIF5A','EIF6')

# Long format
df.long <- df.int %>% pivot_longer(all_of(int.genes), names_to = 'Gene', values_to = 'Expression')
 
# Violin plot
p_violin <- df.long %>%
  # Plot genes of interest
  filter(Gene %in% c('SOX4','EIF5A')) %>%
  filter(subclone != 'Other') %>%
  mutate(Gene = factor(Gene, levels = int.genes)) %>%
  # Make genes italic
  mutate(Gene <- paste0('*', Gene, '*')) %>%
  ggplot(aes(x = stage, y = Expression)) +
  ggrastr::geom_quasirandom_rast(aes(color = stage), size = 0.5) +
  scale_color_manual(values = col_pair,
                     breaks = names(col_pair)) +
  stat_summary(fun.data = "mean_cl_boot", geom = "pointrange", colour = "black", size = 0.5) + # 95% confidence interval for the mean
  ggh4x::facet_nested(. ~ patient + Gene, nest_line = element_line(linetype = 2), solo_line = F, resect = unit(0.5, "cm"),
                      scales = 'free_x', space = 'free_x') +
  theme_pubr(legend = 'none') +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        ggh4x.facet.nestline = element_line(colour = "darkgrey"),
        strip.background = element_blank(),
        strip.text = element_text(face = "italic")
  )

p_violin

cowplot::save_plot(file = './figures/Figure6f_P5_violinPlot_relapse_genes.pdf', plot = p_violin, base_width = 1.7, ncol = length(int.genes))
  
# UMAP
p_umap <- df.long %>%
  # Plot genes of interest
  filter(Gene %in% c('SOX4','EIF5A')) %>%
  mutate(Gene = factor(Gene, levels = c('SOX4','EIF5A'))) %>%
  # Make genes italic
  mutate(Gene <- paste0('*', Gene, '*')) %>%
  split(.$Gene) %>%
  purrr::map(~ ggplot(., aes(wnnUMAP_1, wnnUMAP_2)) +
        ggrastr::geom_point_rast(data = . %>% filter(is.na(Expression)), color = 'lightgrey',
                                 size = 0.25, stroke = 0, raster.dpi = 700) +
        ggrastr::geom_point_rast(data = . %>% filter(!is.na(Expression)), aes(color = Expression),
                                 size = 0.5, stroke = 0, raster.dpi = 700) +
        scale_colour_gradientn(colors = viridis::inferno(20)[1:19]) +
        facet_grid(.~Gene) +
        theme_minimal() + 
        theme(legend.text = element_text(size = 12),
              legend.position = 'bottom',
              #legend.title = element_blank(), 
              axis.text=element_text(size=12),
              axis.title=element_text(size=14),
              axis.line=element_blank(),
              axis.text.x=element_blank(),
              axis.text.y=element_blank(),
              axis.title.x=element_blank(),
              axis.title.y=element_blank(),
              axis.ticks=element_blank(),
              panel.grid = element_blank(),
              strip.text = element_text(face = "italic") #ggtext::element_markdown() 
        ) +
        guides(fill = guide_legend(override.aes = list(size=3), title = "Expression")) +
        coord_fixed() +
        labs(x = "UMAP 1", y = "UMAP 2")) %>%
   cowplot::plot_grid(plotlist = ., ncol = 2)
  
p_umap

cowplot::save_plot(file = './figures/Figure6f_P5_UMAP_relapse_genes.pdf', plot = p_umap, base_width = 2.5, ncol = 2)
  
```
## Plot Figure 6g lollipop plots

Perform pathway analysis using DEGs between diagnosis and relapse. 

*Supplementary Table 10 contains DEGs for all AML cells between diagnosis and refractory disease (Supplementary Table 10-P5) as well as DEGs for subclone-specific differences between the two stages (Supplementary Table 10-P5_SC1). The script below can be used to perform pathway analysis for all the indicated comparisons and plot the results.*

```{r}
markers.df <- readxl::read_excel('./data/Supplementary_tables.xlsx', skip = 2, sheet = 'Supplementary Table 10-P5')

# Top markers at diagnosis
top.markers.de.diagnosis <- as.data.frame(markers.df) %>% filter(FDR < 0.05 & summary.logFC > 0.25 & Group == 'Diagnosis')
dim(top.markers.de.diagnosis)

# Top markers at relapse
top.markers.de.relapse <- as.data.frame(markers.df) %>% filter(FDR < 0.05 & summary.logFC > 0.25 & Group %in% c('Refractory', 'Relapse'))
dim(top.markers.de.relapse)

# Pathway analysis using "Molecular Signatures Database (MSigDB)" gene sets
m_df = msigdbr(species = "Homo sapiens", category = "H")

m_t2g = m_df %>% dplyr::select(gs_name, gene_symbol) %>% as.data.frame()

# Marker over-presentation analysis
resHallmark1 <- clusterProfiler::enricher(top.markers.de.diagnosis$Gene, minGSSize = 10, TERM2GENE = m_t2g, pAdjustMethod = 'BH', pvalueCutoff = 0.05)
resHallmark2 <- clusterProfiler::enricher(top.markers.de.relapse$Gene, minGSSize = 10, TERM2GENE = m_t2g, pAdjustMethod = 'BH', pvalueCutoff = 0.05)

df.resHallmark1 <- resHallmark1@result
df.resHallmark2 <- resHallmark2@result

# Add group column and combine
df.resHallmark1$Group <- 'Diagnosis'
df.resHallmark2$Group <- 'Relapse'
df.resHallmark <- rbind(df.resHallmark1, df.resHallmark2)

# Clean-up naming
df.resHallmark$ID <- str_replace(df.resHallmark$ID, 'HALLMARK_', '')
df.resHallmark$ID <- str_replace_all(df.resHallmark$ID, '_', ' ')
df.resHallmark$ID <- str_to_title(df.resHallmark$ID)
```

Plot lollipop plots.
```{r}
# Focus on gene sets with a potential importance in AML
set.remove <- c('Estrogen Response Early', 'Estrogen Response Late', 'Androgen Response', 'Cholesterol Homeostasis', 'Epithelial Mesenchymal Transition', 'Angiogenesis', 'Pancreas Beta Cells', 'Myogenesis', 'Xenobiotic Metabolism', 'Apical Surface', 'Bile Acid Metabolism', 'Adipogenesis')

p_lolli <- df.resHallmark %>% 
  mutate(Group = factor(Group, levels = c('Diagnosis','Relapse'))) %>%
  filter(!ID %in% set.remove) %>%
  filter(p.adjust < 0.2) %>%
  # Reorder ID based on p.adjust
  mutate(ID_group = fct_reorder(paste0(ID, "-", Group), -log10(p.adjust))) %>%
  ggplot(aes(y = -log10(p.adjust), x = ID_group)) +
  # Add line to FDR = 0.1
  geom_hline(yintercept = -log10(0.05), linetype = 'dashed', color = 'darkgrey') +
  # Add data
  geom_segment(aes(yend = 0, xend = ID_group)) +
  geom_point(aes(size = -log10(p.adjust), color = Group)) +
  scale_color_manual(values = col_pair,
                     breaks = names(col_pair)) +
  facet_grid(Group ~ ., scales = 'free', space = 'free', drop = TRUE) +
  labs(x = NULL,
       y = '-log10(FDR)',
       title = 'Enriched pathways') +
  labs(fill="") +
  coord_flip(clip = 'off') +
  scale_y_continuous(expand = expansion(add = c(0, 0.1)),
                     position = "left", limits = c(0, -log10(min(df.resHallmark$p.adjust)) + 0.5)) +
  scale_x_discrete(labels = function(x) str_remove(x, "-.+")) +
  theme_pubr(legend = 'right') +
  theme(strip.background = element_blank(),
        strip.text.y = element_blank(),
        legend.text = element_text(size = 12),
        plot.title = element_text(vjust = 0.5)) +
  guides(size = guide_legend(title = '-log10(FDR)'), color = guide_legend(override.aes = list(size = 3)))

p_lolli

cowplot::save_plot("./figures/Figure6g_P5_ORA_Lollipop.pdf", p_lolli, base_width = 5.5)
```

